Clustering, profiling and plots

### 2024-06-03 data/03-clustering-and-profiling.qmd
Draft using SOM method

---
title: "Clustering"
author: "Antonio Paez & Anastasia Soukhov"
format: html
---

## Quarto

Lets begin by loading the packages used for this:
```{r message = FALSE}
library(AppliedPredictiveModeling) # Functions and Data Sets for 'Applied Predictive Modeling'
library(caret) # Classification and Regression Training
library(ggparty)
library(ggpubr)
library(party)
library(plotly) # Create Interactive Web Graphics via 'plotly.js'
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(skimr)
#library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(tidyverse) # Easily Install and Load the 'Tidyverse'

# library(formattable) # Create 'Formattable' Data Structures
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
# library(plotly) # Create Interactive Web Graphics via 'plotly.js'
# library(sf) # Simple Features for R
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
# library(tidyverse) # Easily Install and Load the 'Tidyverse'
# library(tmap) # Thematic Maps
```

Read od table:
```{r}
load("access_15.rda")
```

Aggregate by Care Category:
```{r}
access_15_aggregated <- access_15 |>
  group_by(from_id,
           Care_Category) |>
  summarize(copp_isc = sum(copp_isc),
            .groups = "drop")
```

@@

Now this table can be used for the clustering analysis.

Given these normalized attributes, a SOM can be trained using the function `trainSOM` from `SOMbrero`. Only one argument is needed, the input data `x.data`. Other arguments can be assigned by default. In what follows the dimension of the map is given as 5-by-5, and I request that 100 intermediate steps be saved. 
```{r}
set.seed(593)
# run the SOM algorithm 
nrows <- 5
ncols <- 5
access.som <- trainSOM(x.data = access_15_wide |>
                       # Remove the non-normalized accessibility scores
                       select(-c(from_id, starts_with("total_"))),
                     dimension=c(nrows, ncols),
                     nb.save = 100,
                     radius.type = "letremy",
                     verbose = TRUE)
```

@@

The lower left corner of the map concentrates monsters with high speed, high special attack, and high special attack. the upper left corner concentrates high defense and high special defense. High hit points are towards the center of the map.

Two additional way of visualizing the characteristics of the clusters are as follows.

@@

Super clusters:
```{r}
access.sc <- superClass(access.som,
                        k = 6)
summary(access.sc)
```

This is a dendogram with the super-clusters:
```{r}
plot(access.sc, 
     plot.var = TRUE)
```

And the map with the superclusters:
```{r}
plot(access.sc, 
     type = "grid", 
     plot.legend = TRUE)
```

@@

Select data for classification:
```{r}
df <- access_15_wide |>
  select(-c(cluster,
            from_id, 
            starts_with("prop"), 
            starts_with("norm"))) |>
  rename(Dependent = total_Dependent_centric,
         Grocery = total_Grocery_centric,
         Errand = total_Errand_centric,
         Health = total_Health_centric) |>
  mutate(scluster = factor(scluster),
         div = round(div, digits = 3))
```

Train a decision tree using the data:
```{r}
classmod.som <- rpart(scluster ~ ., 
                      data = df)
```

@@

Frequency of parcels in the super clusters (look at the data of the first node):
```{r}
freq_sc <- ggplot(classmod.sompar[1]$data) +
  geom_bar(aes(x = "", 
               fill = scluster)) +
  scale_fill_discrete(name = "super cluster") +
  theme_void()

@@

Extract the legend:
```{r}
leg <- ggpubr::get_legend(freq_sc,
                          position = "bottom")

leg <- as_ggplot(leg)

leg
```

Make a prettier plot:
```{r}
party_plot <- ggparty(classmod.sompar) +
  geom_edge() +
  geom_edge_label(size = 1.5) +
  geom_node_splitvar(size = 2) +
  # pass list to gglist containing all ggplot components we want to plot for each
  # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
```

Plot the tree:
```{r}
party_plot
```

>>>>

---
title: "Clustering, profiling and plots"
author: "Antonio Paez & Anastasia Soukhov"
format: html
---

## Setting up table

Lets begin by loading the packages used for this:
```{r message = FALSE}
library(AppliedPredictiveModeling) # Functions and Data Sets for 'Applied Predictive Modeling'
library(caret) # Classification and Regression Training
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(party) # A Laboratory for Recursive Partytioning
library(plotly) # Create Interactive Web Graphics via 'plotly.js'
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(skimr) # Compact and Flexible Summaries of Data
library(sf) # Simple Features for R
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap)
library(here)
library(parsnip)
```

Read od table and the census:
```{r}
load("access_15.rda")
load("HAM_census_21.rda")
```

Loading boundaries for mapping:
```{r}
Community_Boundaries <- st_read("data-raw/boundaries/Community_Boundaries.shp") |> mutate(COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton-Central", COMMUNITY_))|> st_transform(crs=4326)
City_Boundary <- st_read("data-raw/boundaries/City_Boundary.shp") |> st_transform(crs=4326)
hydro_p_LakeOntario <- st_read("data-raw/boundaries/hydro_p_LakeOntario.shp") |> st_transform(crs=4326)
ham_bay <- st_read("data-raw/boundaries/Waterbodies.shp") |> st_transform(crs=4326) |> filter(FEATURE_TY == "Lake")
ham_bay_cropped<-st_crop(ham_bay$geometry, Community_Boundaries$geometry)
hydro_p_LakeOntario_cropped<-st_crop(hydro_p_LakeOntario$geometry, Community_Boundaries$geometry)
```

Aggregate by Care Category (dependent, errand, health, and grocery centric):
```{r}
access_15_aggregated <- access_15 |>
  group_by(from_id,
           Care_Category) |>
  summarize(GeoUID = first(GeoUID),
            copp_isc = sum(copp_isc),
            .groups = "drop") |> 
  left_join(HAM_census_21 |> # and dssociate with census data and drop parcels in DAs with 0 population:
              st_drop_geometry(),
            by = "GeoUID")

access_15_aggregated <- access_15_aggregated |> filter(Population >= 0 | !is.na(Population))
```

@@

## Plotting population descriptions, DA

For plotting, create a spatial DA sf object that summarizes these inputs:
```{r}
access_15_DA <- left_join(access_15_wide, 
                          access_15_aggregated |> group_by(from_id) |>
  summarise(GeoUID = first(GeoUID),
            tot_copp_isc = first(tot_copp_isc),
            Population = first(Population),
            LICO = first(`v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`),
            Income_bottom_n = first( `v_CA21_1103: In bottom half of the distribution`),
            Income_top_n = first(`v_CA21_1121: In top half of the distribution`),
            Age_0to14_n = first(`v_CA21_11: 0 to 14 years`),
            Age_15to64_n = first(`v_CA21_68: 15 to 64 years`),   
            Age_65plus_n = first(`v_CA21_251: 65 years and over`),
            One_parent_families_n = first(`v_CA21_507: Total one-parent families`),
            Couple_families_n = first(`v_CA21_500: Total couple families`)),
            by=c("from_id")) |>
  group_by(GeoUID)|>
  summarise(Population = first(Population),
            parcel_count = n(),
            LICO = first(LICO),
            Income_bottom_n = first(Income_bottom_n),
            Income_top_n = first(Income_top_n),
            Age_0to14_n = first(Age_0to14_n),
            Age_15to64_n = first(Age_15to64_n),   
            Age_65plus_n = first(Age_65plus_n),
            One_parent_families_n = first(One_parent_families_n),
            Couple_families_n = first(Couple_families_n),

            med_tot_copp_isc = median(tot_copp_isc),
            med_div = median(div),
            med_dep_copp = median(total_Dependent_centric),
            med_err_copp = median(total_Errand_centric),
            med_groc_copp = median(total_Grocery_centric),
            med_health_copp = median(total_Health_centric)) |>
  left_join(HAM_census_21 |> select(c(GeoUID)), by=c("GeoUID")) |> st_as_sf() |> st_make_valid()

access_15_DA <- access_15_DA |> mutate(Populationparcel = Population/parcel_count,
                                       Income_bottom = Income_bottom_n / (Income_bottom_n + Income_top_n),
                                       Age_0to14 = Age_0to14_n / (Age_0to14_n + Age_15to64_n + Age_65plus_n),
                                       One_parent_families = One_parent_families_n / (One_parent_families_n + Couple_families_n))

access_15_DA$parcel_count |> sum() #a check
```

```{r, eval=FALSE}
parcels_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("parcel_count",
              border.alpha = 0.2,
              palette = "Reds",style = "quantile", n=4,
              title = "Count of residential parcels",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5,legend.title.size = 0.7,  legend.bg.color = "white", legend.bg.alpha = 0.4)

pop_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("Population",
              border.alpha = 0.2,
              palette = "Greens", style = "quantile", n=4,
              title = "Population",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5,legend.title.size = 0.7,  legend.bg.color = "white", legend.bg.alpha = 0.4)

poprate_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("Populationparcel",
              border.alpha = 0.2,
              palette = "Purples", style = "quantile", n=4,
              title = "Population per residential parcel rate",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5,legend.title.size = 0.7,  legend.bg.color = "white", legend.bg.alpha = 0.4)

pop_descriptive_plots_1 <- tmap_arrange(parcels_DA_plot, pop_DA_plot, poprate_DA_plot, nrow=3)
tmap_save(pop_descriptive_plots_1,
          file= file.path(here("figures/pop_descriptive_plots_1.png")) , asp = 1/0.6,
          dpi=300)
```

```{r, eval=FALSE}
INCOME_B_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("Income_bottom",
              border.alpha = 0.2,
              palette = "Greens", style = "quantile", n=4,
              title = "% hld below AT median income",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5, legend.title.size = 0.7, legend.bg.color = "white", legend.bg.alpha = 0.4)

AGE014_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("Age_0to14",
              border.alpha = 0.2,
              palette = "Oranges", style = "quantile", n=4,
              title = "% population aged 0 to 14",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5, legend.title.size = 0.7, legend.bg.color = "white", legend.bg.alpha = 0.4)

ONE_PARENT_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("One_parent_families",
              border.alpha = 0.2,
              palette = "Reds", style = "quantile", n=4,
              title = "% one parent hlds",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5, legend.title.size = 0.7, legend.bg.color = "white", legend.bg.alpha = 0.4)

LICOAT_DA_plot <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("LICO",
              border.alpha = 0.2,
              palette = "Purples", style = "quantile", n=4,
              title = "Prevalance of household LICO-AT",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5, legend.title.size = 0.7, legend.bg.color = "white", legend.bg.alpha = 0.4)

pop_descriptive_plots_2 <- tmap_arrange(INCOME_B_DA_plot,AGE014_DA_plot, ONE_PARENT_DA_plot, LICOAT_DA_plot, ncol=2)

tmap_save(pop_descriptive_plots_2,
          file= file.path(here("figures/pop_descriptive_plots_2.png")) , asp = 1/1,
          dpi=300)
```

## Plotting accessibility and diversity index, DA

```{r, eval=FALSE}
all_cats_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_tot_copp_isc",
              border.alpha = 0.2,
              palette = c("darkred", "orangered", "forestgreen", "lightgreen"),
              style = "quantile", n=4,
              #breaks = c(0,5,14,27,108),
              title = "The number of care destinations within a 15 min. walk for a median parcel",
              label = c("[0 to 5) Not caring",
                        "[5 to 14) Somewhat caring",
                        "[14 to 27) Caring",
                        "[27 to 108] Very caring"),  #these are for quantiles, the clusters p50 seem to roughly follows these.
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("right", "top"), size=1.0)+
  tm_layout(bg.color = "grey",
            panel.label.bg.color = "white",
            panel.label.color = "black",
            legend.outside = TRUE,
            legend.position = c("center", "bottom"),
            legend.outside.position = "bottom",
            legend.outside.size = 0.2,
            legend.width = 12
            # legend.title.size = 0.7,
            # legend.text.size = 0.6
            )
tmap_save(all_cats_copp_plot, file= file.path(here("figures/all_cats_copp_plot.png")), dpi=300)
```

```{r, eval=FALSE}
dep_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_dep_copp",
              border.alpha = 0.2,
              style = "quantile", n=4,
              palette = "Oranges",
              title = "Dependent-centric",) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))

health_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_health_copp",
              border.alpha = 0.2,
              style = "quantile", n=4,
              palette = "Purples",
              title = "Health-centric",) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))

groc_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_groc_copp",
              border.alpha = 0.2,
              style = "quantile", n=4,
              palette = "Greens",
              title = "Grocery-centric",) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))

err_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_err_copp",
              border.alpha = 0.2,
              style = "quantile", n=4,
              palette = "Reds",
              title = "Errand-centric",) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))

percats_copp_plot <- tmap_arrange(dep_copp_plot, err_copp_plot, groc_copp_plot, health_copp_plot, nrow=2)

tmap_save(percats_copp_plot, 
          file= file.path(here("figures/percats_copp_plot.png")),
          dpi=300)
```

```{r}
# very caring (mean 4 to 18.6; total p50 46) - very complete (div=0.9; p50 0.90) (Cluster 2)
# caring (mean 2.1 to 9; total p50 18) -  complete (div=0.81; p50 0.82) (Cluster 1)
# somewhat caring (0.2 to 6; total p50 13) - complete (div=0.71; p50 0.74) (Cluster 4)
# somewhat caring (0.2 to 6; total p50 7) - not complete (div=0.36; p50 0.40) (Cluster 3)
# not caring (0 to 4.7; total p50 3) - not complete (div=0.02; p50 0) (Cluster 5)
```

```{r, eval=FALSE}
all_cats_entropy_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("med_div",
              border.alpha = 0.2,
              style = "quantile", n=3, #are the quantiles, but are also similar to the clustered profiles
              label = c("[0 to 0.49) Not complete", "[0.49 to 0.85) Complete", "[0.85 to 1] Very complete"),
              palette = "Greys",
              title = "Diversity of care destinations within a 15-minute walk for a median parcel",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("right", "top"), size=1.0)+
  tm_layout(bg.color = "grey",
            panel.label.bg.color = "white",
            panel.label.color = "black",
            legend.outside = TRUE,
            legend.position = c("center", "bottom"),
            legend.outside.position = "bottom",
            legend.outside.size = 0.2,
            legend.width = 12)

tmap_save(all_cats_entropy_copp_plot, file= file.path(here("figures/all_cats_entropy_copp_plot.png")), dpi=300)
```

## Creating clusters (from parcels)

Now access_15_wide can be used for the clustering analysis.

Given these normalized attributes, a SOM can be trained using the function `trainSOM` from `SOMbrero`. Only one argument is needed, the input data `x.data`. Other arguments can be assigned by default. In what follows the dimension of the map is given as 5-by-5, and I request that 100 intermediate steps be saved. 
```{r}
set.seed(5935)
# run the SOM algorithm 
nrows <- 5
ncols <- 5
access.som <- trainSOM(x.data = access_15_wide |>
                       # Remove the non-normalized accessibility scores
                       select(-c(from_id, starts_with("total_"))),
                     dimension=c(nrows, ncols),
                     nb.save = 100,
                     radius.type = "letremy",
                     verbose = TRUE)
```

@@

The lower right corner of the map concentrates monsters has low levels of all variables. The high concentration of all variables is found in the upper left corner. Two additional way of visualizing the characteristics of the clusters are as follows.

@@

Super clusters:
```{r}
access.sc <- superClass(access.som,
                        k = 5)
summary(access.sc)
```
The combined dendogram and grid-map of the super-clusters. Turned off the variance explained plot:
```{r, eval=FALSE}
png(file= file.path(here("figures/access_sc_plot_1.png")), width= 2404, height= 1600, units="px",  res=300)
plot(access.sc, 
     type= c("dendrogram"),
     plot.var = FALSE,
     show.names = TRUE,
     main = " ",
     ylab="Dissimilarity Index",
     xlab= " ")
dev.off()
```

```{r, eval=FALSE}
png(file= file.path(here("figures/access_sc_plot_2.png")), width= 2404, height= 1600, units="px",  res=300)
plot(access.sc,
     type = "grid", 
     plot.var = FALSE,
     main = " ",
     plot.legend = TRUE)
dev.off()
```

@@

Select data for classification:
```{r}
df <- access_15_wide |>
  select(-c(cluster,
            starts_with("prop"), 
            starts_with("norm"))) |>
  rename(Dependent = total_Dependent_centric,
         Grocery = total_Grocery_centric,
         Errand = total_Errand_centric,
         Health = total_Health_centric) |>
  mutate(scluster = factor(scluster),
         div = round(div, digits = 3))

df <- df |> 
  mutate(Total_copp = Dependent + Grocery + Errand + Health)
```

Train a decision tree using the data:
```{r}
classmod.som <- rpart(data = df,
                      formula = scluster ~ (Dependent + Grocery + Errand + Health + div))

# Predict the labels and add them to the dataset
predicted_labels <- predict(classmod.som, df) |> as.data.frame()

df$SC1 <- predicted_labels$`1`
df$SC2 <- predicted_labels$`2`
df$SC3 <- predicted_labels$`3`
df$SC4 <- predicted_labels$`4`
df$SC5 <- predicted_labels$`5`

# df <- df |> mutate(max_label_val = pmax(predicted_labels$`1`, predicted_labels$`2`, predicted_labels$`3`, predicted_labels$`4`, predicted_labels$`5`),
#                    max_label = case_when(max_label_val == predicted_labels$`1` ~ "SC1",
#                                          max_label_val == predicted_labels$`2` ~ "SC2",
#                                          max_label_val == predicted_labels$`3` ~ "SC3",
#                                          max_label_val == predicted_labels$`4` ~ "SC4",
#                                          max_label_val == predicted_labels$`5` ~ "SC5"))

df_DA <- df |> left_join(access_15 |> dplyr::select(c("from_id","GeoUID")), by = "from_id") |> 
  group_by(GeoUID) |> 
  summarise(
    SC1 = median(SC1),
    SC2 = median(SC2),
    SC3 = median(SC3),
    SC4 = median(SC4),
    SC5 = median(SC5),
    max_label_val = pmax(SC1, SC2, SC3, SC4, SC5),
            med_tot_copp_isc = median(Total_copp),
            med_div = median(div),
            med_dep_copp = median(Dependent),
            med_err_copp = median(Errand),
            med_groc_copp = median(Grocery),
            med_health_copp = median(Health)) |>
  mutate(max_label = case_when(max_label_val == SC1 ~ "b. Caring - complete (SC1)",
                               max_label_val == SC2 ~ "a. Very caring - very complete (SC2)",
                               max_label_val == SC3 ~ "d. Somewhat caring - not complete (SC3)",
                               max_label_val == SC4 ~ "c. Somewhat caring - complete (SC4)",
                               max_label_val == SC5 ~ "d. not caring - not complete (SC5)")) |>
  left_join(HAM_census_21 |> select(c(GeoUID)), by=c("GeoUID")) |> st_as_sf() |> st_make_valid() 
```
```{r, eval=FALSE}
total_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(df_DA) +
  tm_polygons("med_tot_copp_isc",
              border.alpha = 0.2,
              style = "quantile", n=4,
              palette = "Purples",
              title = "Total access to care destinations",) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))

SClabel_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(df_DA) +
  tm_polygons("max_label",
              border.alpha = 0.2,
              palette = c("darkolivegreen","red", "dodgerblue", "darkorchid","darkgreen"),
              labels = (c("Very caring - very complete (SC2)","Caring - Complete (SC1)",
              "Somewhat caring - complete (SC4)","Somewhat caring - somewhat complete (SC5)", 
              "Somewhat caring - somewhat complete (SC3)")),
              title = "Prevelant supercluster") +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))
SClabel_copp_plot
SCLab_and_total_copp_plot <- tmap_arrange(SClabel_copp_plot, total_copp_plot, nrow=2)

tmap_save(SCLab_and_total_copp_plot, 
          file= file.path(here("figures/SCLab_and_total_copp_plot.png")),
          dpi=300)
```

@@

Frequency of parcels in the super clusters:
```{r}
freq_sc <- ggplot(access_15_wide |> 
         mutate(scluster = factor(scluster))) +
  geom_bar(aes(x = "", 
               group = scluster, 
               fill = scluster)) +
  scale_fill_discrete(name = "Supercluster") #+
  #theme_void() +
  #guides(fill = guide_legend(nrow = 1)) +
  #theme(legend.position = "bottom")

@@

Extract the legend:
```{r}
legd <- cowplot::get_legend(freq_sc)

legd
```

Make a prettier plot:
```{r}
party_plot <- ggparty(classmod.sompar) +
  geom_edge() +
  geom_edge_label(size = 1.6) +
  geom_node_label(aes(label = splitvar),
                  id = "inner", 
                  size = 2, nudge_y = -0.022) +
  # pass list to gglist containing all ggplot components we want to plot for each
  # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
party_plot
```

Plot the tree:
```{r}
png(file= file.path(here("figures/decision_tree_profiles_sc.png")),width= 2404, height= 1600, units="px",  res=300)
plot_grid(party_plot, legd, nrow = 1, rel_widths = c(1, 0.2))
dev.off()
```

Cluster 6 is high in suckitude: low in diversity, with 4.6 Dependent Centric destinations on average, and typically no other destinations: zero Errand centric destinations, and only about 3.5% of parcels have access to Grocery-centric destinations.
```{r}
access_15_wide |>
  select(scluster, starts_with("total_"), div) |>
  rename_with(.fn = ~str_replace(.x, "total_", "")) |>
  filter(scluster == 1) |>
  skim()
```

What I've deduced from these superclusters:

very caring (mean 4 to 18.6; total p50 46) - very complete (div=0.9; p50 0.90) (Cluster 2)
caring (mean 2.1 to 9; total p50 18) - complete (div=0.81; p50 0.82) (Cluster 1)
somewhat caring (0.2 to 6; total p50 13) - complete (div=0.71; p50 0.74) (Cluster 4)
somewhat caring (0.2 to 6; total p50 7) - not complete (div=0.36; p50 0.40) (Cluster 3)
not caring (0 to 4.7; total p50 3) - not complete (div=0.02; p50 0) (Cluster 5)

## Demographic profile of clusters
